Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2021-11-29 Manuscript/Manuscript.Rmd
Feedback on figures (md pattern, conditinal distributions, distributions after imputation)

```{r patterns, echo=FALSE}
# monotone vs non-monotone
patterns <- rbind(
  tibble(
    pat = rep("systematic", 5),
    rownr = 5:1,
    # obs = rep("observed", 5),
    cluster = rep("observed", 5),
    X1 = rep("observed", 5),
    X2 = c(rep("observed", 3), rep("missing", 2)),
    X3 = c(rep("missing", 3),  rep("observed", 2))
  ) %>% pivot_longer(cols = c(cluster, X1, X2, X3)),
  tibble(
    pat = rep("sporadic", 5),
    rownr = 5:1,
    # obs = rep("observed", 5),
    cluster = rep("observed", 5),
    X1 = c("observed", "missing", rep("observed", 3)),
    X2 = c(rep("observed", 2), rep("missing", 3)),
    X3 = c(rep("missing", 3),  rep("observed", 2))
  ) %>% pivot_longer(cols = c(cluster, X1, X2, X3))
) %>%
  mutate(across(everything(), as.factor),
         pat = factor(pat, levels = c("systematic", "sporadic"), labels = c("Systematic missingness", "Sporadic missingness"), ordered = TRUE),
         text = rep(c(rep(c("1","", "", ""), 3), rep(c("2","", "", ""), 2)),2))
# plot
ggplot(patterns, aes(
  x = name,
  y = rownr,
  color = value,
  type = pat,
  width = 0.8, 
  height = 0.8,
  label = text
)) +
  geom_tile(fill = "white", size = 2) +
  geom_text() +
  scale_x_discrete(position = "top") +
  scale_y_discrete(labels = 5:1) +
  scale_color_manual(values = plot_col, name = "Datapoint") +
  facet_wrap( ~ pat, ncol = 2, strip.position = "bottom") +
  theme_minimal() +
  theme(
    legend.position = "none", 
    panel.grid.major = element_blank()) +
  labs(x = "", y = "Observation (e.g., patient)")
# TODO: add patient nr and cluster
# TODO: add grid.lines = "none"

@@

```{r impact, echo=FALSE}
# load data
data("impact")
# # descriptive statistics
# by(impact, impact$name, summary) 
# psych::describe(impact)[,c(2:5,8:9)]
# missingness
plot_md_pat(impact)
# TODO: rotate variable labels
```
-> Why are there no missings? According to the [vignette](https://cran.r-project.org/web/packages/metamisc/metamisc.pdf), the data is already imputed (Steyerberg et al, 2008).

@@

- Does the missing data of `popular` depend on `popteach`? One could for example check this by making a histogram of `popteach` separately for the pupils with known popularity and missing popularity. And the other way around: Does the missingness in teacher popularity depend on pupil popularity?

```{r pop_dist, echo=FALSE, message=FALSE, warning=FALSE}
# distributions
plot_NA_cond(pop, x = "popteach", z = "popular", bins = 1)
plot_NA_cond(pop, x = "popular", z = "popteach", bins = 1)
# TODO: check if up-side-down plot works
# or maybe overlapping densities
# TODO: check smoothing in geom_density function and make it the inverse of the sample size
# TODO: add functions to mice
```

-  We do see that the histogram for the missing `popular` is further to the right than the histogram for observed `popular`. This would indicate a right-tailed MAR missingness. In fact this is exactly what happens, because we created the missingness in these data ourselves. But we made it observable by examining the relations between the missingness in popular and the observed data in `popteach`. There is also a dependency between the missingness in teacher popularity and pupil popularity. The relation seems to be right-tailed.

>>>>

```{r patterns, echo=FALSE}
# monotone vs non-monotone
patterns <- rbind(
  tibble(
    pat = rep("systematic", 5),
    rownr = 5:1,
    # obs = rep("observed", 5),
    cluster = rep("observed", 5),
    X1 = rep("observed", 5),
    X2 = c(rep("observed", 3), rep("missing", 2)),
    X3 = c(rep("missing", 3),  rep("observed", 2))
  ) %>% pivot_longer(cols = c(cluster, X1, X2, X3)),
  tibble(
    pat = rep("sporadic", 5),
    rownr = 5:1,
    # obs = rep("observed", 5),
    cluster = rep("observed", 5),
    X1 = c("observed", "missing", rep("observed", 3)),
    X2 = c(rep("observed", 2), rep("missing", 3)),
    X3 = c(rep("missing", 3),  rep("observed", 2))
  ) %>% pivot_longer(cols = c(cluster, X1, X2, X3))
) %>%
  mutate(across(everything(), as.factor),
         pat = factor(pat, levels = c("systematic", "sporadic"), labels = c("Systematic missingness", "Sporadic missingness"), ordered = TRUE),
         text = rep(c(rep(c("1","", "", ""), 3), rep(c("2","", "", ""), 2)),2))
# plot
ggplot(patterns, aes(
  x = name,
  y = rownr,
  color = value,
  type = pat,
  width = 0.8, 
  height = 0.8,
  label = text
)) +
  geom_tile(fill = "white", size = 2) +
  geom_text() +
  scale_x_discrete(position = "top") +
  scale_y_discrete(labels = 5:1) +
  scale_color_manual(values = plot_col, name = "Datapoint") +
  facet_wrap( ~ pat, ncol = 2, strip.position = "bottom") +
  theme_minimal() +
  theme(
    legend.position = "none", 
    panel.grid.major = element_blank()) +
  labs(x = "", y = "Observation (e.g., patient)")
# TODO: make it one plot with sporadic and systematic
# TODO: participant/response/*sample units*/records instead of obs on the y axis

@@

```{r impact, echo=FALSE}
# load data
data("impact")
# # descriptive statistics
# by(impact, impact$name, summary) 
# psych::describe(impact)[,c(2:5,8:9)]
# missingness
plot_md_pat(impact)
# TODO: rotate variable labels

# TODO: check md pattern in clusters, should differ, otherwise create new missingness
# MAYBE: think about nr of clusters: if there's too many, faceting is no option -> some distribution across the clusters instead
# TODO: e-mail Joop for DGM or look at adrew and jennifer book for data
# sim.multi from psych
```
-> Why are there no missings? According to the [vignette](https://cran.r-project.org/web/packages/metamisc/metamisc.pdf), the data is already imputed (Steyerberg et al, 2008).

@@

- Does the missing data of `popular` depend on `popteach`? Does the missingness in teacher popularity depend on pupil popularity? -> Check this by making a histogram of `popteach` separately for the pupils with known popularity and missing popularity, and the other way around.

```{r pop_dist, echo=FALSE, message=FALSE, warning=FALSE}
# distributions
# plot_NA_cond(pop, x = "popteach", z = "popular", bins = 1)
# plot_NA_cond(pop, x = "popular", z = "popteach", bins = 1)
# TODO: check if up-side-down plot works
# or maybe overlapping densities
# TODO: check smoothing in geom_density function and make it the inverse of the sample size
# TODO: add functions to mice

miss_col <- mice:::mdc(1:2) %>% setNames(c("FALSE", "TRUE"))

# x = "popular"
# z = "popteach"
plot_conditional <- function(data, x, z, cluster, ...){
  ggplot(data, aes(x = get(x), color = !is.na(get(z)))) + # group = get(cluster)
  geom_density(alpha = 0.01) +
  scale_color_manual(values = miss_col, name = z) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = x)
}
plot_conditional(pop, x = "popular", z = "popteach", cluster = "class")
plot_conditional(pop, x = "popteach", z = "popular", cluster = "class")
# TODO: pooteach obs/mis
# TODO: add title
# TODO: add facets for some clusters
# or add propensity score distribution `is.na(popular) ~ .`
# TODO: make it average of cluster dens, not marginal
```

-  We do see that the distribution for the missing `popular` is further to the right than the distribution for observed `popular`. This would indicate a right-tailed MAR missingness. In fact this is exactly what happens, because we created the missingness in these data ourselves. But we made it observable by examining the relations between the missingness in popular and the observed data in `popteach`. There is also a dependency between the missingness in teacher popularity and pupil popularity. The relation seems to be right-tailed as well.

- We can impute the missingness the 'standard' way, ignoring the multilevel structure of the data. This is surely invalid, given the high ICCs, but we'll do it anyways.

- We'll use predictive mean matching to impute the continuous variables (some appear to be somewhat ordinal), and logistic regression to impute the binary variable `sex`. We do not use the observation identifier `pupil` or cluster identifier `class` as predictors to impute other variables.

```{r pop-ignored}
# dry run to get imputation parameters
ini <- mice(pop, maxit = 0)

# extract predictor matrix and adjust
pred <- ini$pred
pred[, c("class", "pupil")] <- 0
pred

# impute the data, ignoring the cluster structure
imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)

# check convergence of the imputation model
plot(imp_ignored)

# compare descriptives before and after imputation
psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
# TODO: add stripplot with boxplot overlay instead of the tables (make pooled one thick on top)
# TODO: pool mean median and sd
# feedback Stef: numbers, continuous statistics such as means, and uncertainty estimates. So we can pool the sd's. And leave out the min and max, because those are not normally distr.
# TODO: add FMI for each of the estimates? at least for the mean

# further inspection of the imputations
densityplot(imp_ignored)

# compare ICCs before and after imputation
ICCs <- data.frame(
  vars = c("popular", "popteach", "texp"), 
  incomplete = c(multilevel::ICC1(aov(popular ~ class, pop)), 
               multilevel::ICC1(aov(popteach ~ class, pop)),
               multilevel::ICC1(aov(texp ~ class, pop))), 
  ignored = c(multilevel::ICC1(aov(popular ~ class, complete(imp_ignored))), 
              multilevel::ICC1(aov(popteach ~ class, complete(imp_ignored))), 
              multilevel::ICC1(aov(texp ~ class, complete(imp_ignored))))
  )
ICCs

- As the original ICCs show, 100% of the variance in `texp` can be attributed to the clustering variable `class`. This tells us that the multilevel structure of the data should be taken into account. If we don't, we'll end up with incorrect imputations, biasing the effect of the clusters towards zero.

- We can also observe that the teacher experience increases slightly after imputation. This is due to the MNAR missingness in `texp`. Higher values for `texp` have a larger probability to be missing. This may not a problem, however, if at least one pupil in each class has teacher experience recorded, we can deductively impute the correct (i.e. true) value for every pupil in the class.

- We'll now use `class` as a predictor to impute all other variables.

```{r pop-predictor}
# adjust the predictor matrix
pred <- ini$pred 
pred[, "pupil"] <- 0
pred

# impute the data, cluster as predictor
imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)

# check logged events
head(imp_predictor$loggedEvents)
## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and popteach.

# check convergence of the imputation model
plot(imp_predictor)

# compare descriptives before and after imputation
psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results

# further inspection of the imputations
densityplot(imp_predictor)

# compare ICCs before and after imputation
ICCs <- ICCs %>% cbind(
           predictor = c(multilevel::ICC1(aov(popular ~ class, complete(imp_predictor))), 
                        multilevel::ICC1(aov(popteach ~ class, complete(imp_predictor))), 
                        multilevel::ICC1(aov(texp ~ class, complete(imp_predictor))))
           )
ICCs
```

- Now, we can clearly see that the imputed values of `texp` are higher than the observed values, which is in line with right-tailed MNAR.

- The ICCs are way more in line with the ICCs in the incomplete data. But this is a quick and dirty way of imputing multilevel data. We *should* be using a multilevel model.

### 2021-11-29 Manuscript/Manuscript.tex
Add steps from mice vignette up-to multilevel models
Feedback on figures (md pattern, conditinal distributions, distributions after imputation)

\begin{itemize}
\item
  leiden85
\item
  GJRM::hiv
\item
  simulating
\item
  IMPACT
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Does the missing data of \texttt{popular} depend on \texttt{popteach}?
  One could for example check this by making a histogram of
  \texttt{popteach} separately for the pupils with known popularity and
  missing popularity. And the other way around: Does the missingness in
  teacher popularity depend on pupil popularity?
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  We do see that the histogram for the missing \texttt{popular} is
  further to the right than the histogram for observed \texttt{popular}.
  This would indicate a right-tailed MAR missingness. In fact this is
  exactly what happens, because we created the missingness in these data
  ourselves. But we made it observable by examining the relations
  between the missingness in popular and the observed data in
  \texttt{popteach}. There is also a dependency between the missingness
  in teacher popularity and pupil popularity. The relation seems to be
  right-tailed.
\end{itemize}

>>>>

\begin{itemize}
\item
  leiden85
\item
  GJRM::hiv (\url{https://rdrr.io/github/egeminiani/GJRM/man/hiv.html})
\item
  simulating
\item
  IMPACT
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Does the missing data of \texttt{popular} depend on \texttt{popteach}?
  Does the missingness in teacher popularity depend on pupil popularity?
  -\textgreater{} Check this by making a histogram of \texttt{popteach}
  separately for the pupils with known popularity and missing
  popularity, and the other way around.
\end{itemize}

@@

\begin{itemize}
\item
  We do see that the distribution for the missing \texttt{popular} is
  further to the right than the distribution for observed
  \texttt{popular}. This would indicate a right-tailed MAR missingness.
  In fact this is exactly what happens, because we created the
  missingness in these data ourselves. But we made it observable by
  examining the relations between the missingness in popular and the
  observed data in \texttt{popteach}. There is also a dependency between
  the missingness in teacher popularity and pupil popularity. The
  relation seems to be right-tailed as well.
\item
  We can impute the missingness the `standard' way, ignoring the
  multilevel structure of the data. This is surely invalid, given the
  high ICCs, but we'll do it anyways.
\item
  We'll use predictive mean matching to impute the continuous variables
  (some appear to be somewhat ordinal), and logistic regression to
  impute the binary variable \texttt{sex}. We do not use the observation
  identifier \texttt{pupil} or cluster identifier \texttt{class} as
  predictors to impute other variables.
\end{itemize}

\begin{CodeChunk}
\begin{CodeInput}
R> # dry run to get imputation parameters
R> ini <- mice(pop, maxit = 0)
R> 
R> # extract predictor matrix and adjust
R> pred <- ini$pred
R> pred[, c("class", "pupil")] <- 0
R> pred 
\end{CodeInput}
\begin{CodeOutput}
         pupil class extrav sex texp popular popteach
pupil        0     0      1   1    1       1        1
class        0     0      1   1    1       1        1
extrav       0     0      0   1    1       1        1
sex          0     0      1   0    1       1        1
texp         0     0      1   1    0       1        1
popular      0     0      1   1    1       0        1
popteach     0     0      1   1    1       1        0
\end{CodeOutput}
\begin{CodeInput}
R> # impute the data, ignoring the cluster structure
R> imp_ignored <- mice(pop, maxit = 10, pred = pred, print = FALSE)
R> 
R> # check convergence of the imputation model
R> plot(imp_ignored)
\end{CodeInput}

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-ignored-1} \end{center}

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-ignored-2} \end{center}

\begin{CodeInput}
R> # compare descriptives before and after imputation
R> psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65   11.0   1  26.0  5.97
class*   2000 50.37   51.0   1 100.0 29.08
extrav   1484  5.31    5.0   1  10.0  1.29
sex*     1504  1.56    2.0   1   2.0  0.50
texp     1024 11.80   12.0   2  25.0  6.26
popular  1490  4.83    4.8   0   9.1  1.34
popteach 1472  4.83    5.0   1  10.0  1.36
\end{CodeOutput}
\begin{CodeInput}
R> psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65     11   1  26.0  5.97
class*   2000 50.37     51   1 100.0 29.08
extrav   2000  5.24      5   1  10.0  1.26
sex*     2000  1.51      2   1   2.0  0.50
texp     2000 12.63     13   2  25.0  6.27
popular  2000  5.01      5   0   9.1  1.37
popteach 2000  5.02      5   1  10.0  1.39
\end{CodeOutput}
\begin{CodeInput}
R> # further inspection of the imputations
R> densityplot(imp_ignored)
\end{CodeInput}

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-ignored-3} \end{center}

\begin{CodeInput}
R> # compare ICCs before and after imputation
R> ICCs <- data.frame(
+   vars = c("popular", "popteach", "texp"), 
+   incomplete = c(multilevel::ICC1(aov(popular ~ class, pop)), 
+                multilevel::ICC1(aov(popteach ~ class, pop)),
+                multilevel::ICC1(aov(texp ~ class, pop))), 
+   ignored = c(multilevel::ICC1(aov(popular ~ class, complete(imp_ignored))), 
+               multilevel::ICC1(aov(popteach ~ class, complete(imp_ignored))), 
+               multilevel::ICC1(aov(texp ~ class, complete(imp_ignored))))
+   )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored
1  popular  0.3280070 0.2743092
2 popteach  0.3138658 0.2479930
3     texp  1.0000000 0.4510366
\end{CodeOutput}
\end{CodeChunk}

\begin{itemize}
\item
  As the original ICCs show, 100\% of the variance in \texttt{texp} can
  be attributed to the clustering variable \texttt{class}. This tells us
  that the multilevel structure of the data should be taken into
  account. If we don't, we'll end up with incorrect imputations, biasing
  the effect of the clusters towards zero.
\item
  We can also observe that the teacher experience increases slightly
  after imputation. This is due to the MNAR missingness in
  \texttt{texp}. Higher values for \texttt{texp} have a larger
  probability to be missing. This may not a problem, however, if at
  least one pupil in each class has teacher experience recorded, we can
  deductively impute the correct (i.e.~true) value for every pupil in
  the class.
\item
  We'll now use \texttt{class} as a predictor to impute all other
  variables.
\end{itemize}

\begin{CodeChunk}
\begin{CodeInput}
R> # adjust the predictor matrix
R> pred <- ini$pred 
R> pred[, "pupil"] <- 0
R> pred
\end{CodeInput}
\begin{CodeOutput}
         pupil class extrav sex texp popular popteach
pupil        0     1      1   1    1       1        1
class        0     0      1   1    1       1        1
extrav       0     1      0   1    1       1        1
sex          0     1      1   0    1       1        1
texp         0     1      1   1    0       1        1
popular      0     1      1   1    1       0        1
popteach     0     1      1   1    1       1        0
\end{CodeOutput}
\begin{CodeInput}
R> # impute the data, cluster as predictor
R> imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)
\end{CodeInput}
\begin{CodeOutput}
Warning: Number of logged events: 190
\end{CodeOutput}
\begin{CodeInput}
R> # check logged events
R> head(imp_predictor$loggedEvents)
\end{CodeInput}
\begin{CodeOutput}
  it im      dep meth  out
1  1  1  popular  pmm texp
2  1  1 popteach  pmm texp
3  1  2  popular  pmm texp
4  1  2 popteach  pmm texp
5  1  3  popular  pmm texp
6  1  3 popteach  pmm texp
\end{CodeOutput}
\begin{CodeInput}
R> ## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and popteach.
R> 
R> # check convergence of the imputation model
R> plot(imp_predictor)
\end{CodeInput}

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-predictor-1} \end{center}

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-predictor-2} \end{center}

\begin{CodeInput}
R> # compare descriptives before and after imputation
R> psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65   11.0   1  26.0  5.97
class*   2000 50.37   51.0   1 100.0 29.08
extrav   1484  5.31    5.0   1  10.0  1.29
sex*     1504  1.56    2.0   1   2.0  0.50
texp     1024 11.80   12.0   2  25.0  6.26
popular  1490  4.83    4.8   0   9.1  1.34
popteach 1472  4.83    5.0   1  10.0  1.36
\end{CodeOutput}
\begin{CodeInput}
R> psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65   11.0   1  26.0  5.97
class*   2000 50.37   51.0   1 100.0 29.08
extrav   2000  5.20    5.0   1  10.0  1.28
sex*     2000  1.52    2.0   1   2.0  0.50
texp     2000 14.26   15.0   2  25.0  6.55
popular  2000  5.05    5.1   0   9.1  1.38
popteach 2000  5.03    5.0   1  10.0  1.39
\end{CodeOutput}
\begin{CodeInput}
R> # further inspection of the imputations
R> densityplot(imp_predictor)
\end{CodeInput}

\begin{center}\includegraphics{Manuscript_files/figure-latex/pop-predictor-3} \end{center}

\begin{CodeInput}
R> # compare ICCs before and after imputation
R> ICCs <- ICCs %>% cbind(
+            predictor = c(multilevel::ICC1(aov(popular ~ class, complete(imp_predictor))), 
+                         multilevel::ICC1(aov(popteach ~ class, complete(imp_predictor))), 
+                         multilevel::ICC1(aov(texp ~ class, complete(imp_predictor))))
+            )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored predictor
1  popular  0.3280070 0.2743092 0.3694899
2 popteach  0.3138658 0.2479930 0.3335612
3     texp  1.0000000 0.4510366 1.0000000
\end{CodeOutput}
\end{CodeChunk}

\begin{itemize}
\item
  Now, we can clearly see that the imputed values of \texttt{texp} are
  higher than the observed values, which is in line with right-tailed
  MNAR.
\item
  The ICCs are way more in line with the ICCs in the incomplete data.
  But this is a quick and dirty way of imputing multilevel data. We
  \emph{should} be using a multilevel model.
\end{itemize}
